Boosting the Accuracy of Finite Difference Schemes via Optimal Time Step Selection and Non-Iterative Defect Correction

### 2011-07-27 papers/OTSPDE/OTSPDE.tex
Added keywords in OTSPDE.tex
Revised OTSPDE paper for resubmission to AMC.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First completed draft: 2008/05/12, Vitamin D, Inc.
% Submission draft: 2008/06/08, Vitamin D, Inc.
% Resubmission draft: 2008/11/16, Vitamin D, Inc.
% Resubmission draft: 2009/05/25, Serendipity Research
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\documentclass[fleqn,12pt,twoside,draft]{article}
\documentclass[fleqn,12pt,twoside]{article}
\usepackage[headings]{espcrc1}
\usepackage{amsmath}
\usepackage{graphicx}
\PassOptionsToPackage{ctagsplt,righttag}{amsmath}
\usepackage{multirow}

@@

\author{Kevin T. Chu\address{Institute of High Performance Computing, A*STAR, Singapore, Singapore}$^,$\address{Vitamin D, Inc., Menlo Park, CA 94025, United States}
}

@@

\begin{abstract}
In this article, we present a simple technique for boosting the order of 
accuracy of finite difference schemes for time dependent 
partial differential equations by optimally selecting the time step used
to advance the numerical solution and adding defect correction terms in a
non-iterative manner.  The power of the technique is its ability to extract 
as much accuracy as possible from existing finite difference schemes with 
minimal additional effort.  Through straightforward numerical analysis 
arguments, we explain the origin of the boost in accuracy and estimate the 
computational cost of the resulting numerical method.  We demonstrate the 
utility of optimal time step (OTS) selection combined with non-iterative 
defect correction (NIDC) on several different types of finite difference
schemes for a wide array of classical linear and semilinear PDEs in one and 
more space dimensions on both regular and irregular domains.  
\end{abstract}

@@

% KEYWORDS
% optimal time step; non-iterative defect correction; finite difference 
% schemes; high-order accurate numerical methods; time dependent PDEs

\section{Introduction}
High-order numerical methods for partial differential equations (PDEs) will 
always be valuable for increasing the computational efficiency of numerical 
simulations.  Thus, it is not at all surprising that a great deal of effort in 
numerical PDEs continues to be focused on the development of high-order 
numerical 
schemes~\cite{spotz_2001,bruger_2005,gibou_2005,ito_2005,shukla_2005,heidenreich_2007,shukla_2007}.  
Typically, high-order accuracy is achieved by constructing
schemes that are formally high-order accurate.  However, high-order 
accuracy can also be obtained by using formally low-order schemes in clever 
ways (\eg Richardson extrapolation and compact finite difference 
schemes~\cite{spotz_2001}).  When possible, the latter approach can be a 
powerful way to boost the accuracy of a numerical method \emph{without} 
introducing too much additional algorithmic (and programming) complexity.

Optimization of numerical parameters (\eg $\dx$ and $\dt$) is an important, 
but relatively uncommon, approach for maximizing the accuracy of numerical 
schemes.  Often, all combinations of numerical parameters are considered 
equally good.  For some problems, however, optimizing the numerical parameters 
can significantly increase the accuracy of the numerical solution.  For 
example, the well-known unit CFL condition, $\dt = \dx/a$, for the first-order 
upwind forward Euler discretization of the linear advection equation 
completely eliminates the numerical error introduced by the 
scheme~\cite{leveque_book_2002}.  Another well-known example is the boost in 
the order of accuracy that results from setting the time step to 
$\dt = \dx^2/6D$ when the 1D diffusion equation is solved using forward Euler 
time integration with the standard second-order central difference 
approximation for the Laplacian.  As a final example, Zhao showed that an 
appropriate choice of the weight factor $\theta$ in the theta-method can 
significantly decrease the error in finite element solutions to the diffusion 
equation~\cite{zhao_2006}.

In this article, we present a simple technique for boosting the order of 
accuracy of finite difference schemes for time dependent PDEs by optimally 
choosing the time step and, when necessary, adding defect correction terms
in a \emph{non-iterative} manner.  This technique is a systematic 
generalization of the ideas underlying some of the examples mentioned above.
Optimal time step (OTS) selection is based on the observation that a carefully 
chosen time step can simultaneously eliminate low-order terms in both the
spatial and temporal the discretization errors.  It combines the well-known 
approach of using knowledge of the structure of the PDE to reduce numerical 
error with the less-common notion that an optimal choice of numerical 
parameters can boost the effectiveness of algorithmically simple and 
computationally low-cost numerical schemes.  For some problems, OTS alone is 
sufficient to allow a formally low-order method to deliver high-order 
accuracy.  However, addition of defect correction terms is generally required 
to eliminate residual terms in the leading-order truncation error.  It is 
important to emphasize that, unlike traditional defect/deferred correction 
methods~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006} 
which use defect correction terms to iteratively refine the solution at 
each time step, we take a non-iterative defect correction (NIDC) approach
that incorporates defect correction terms directly into the original finite 
difference scheme.

As a technique for enhancing numerical methods, OTS-NIDC has several desirable 
features.  Like any technique that leads to high-order methods, OTS-NIDC
produces schemes that greatly reduce the computational cost required to obtain 
a numerical solution of a desired accuracy.  However, its real power comes 
from the fact that it achieves high-order accuracy while being based solely on 
very simple, formally low-order finite difference schemes.  In other words, 
OTS-NIDC allows us to extract as much accuracy as possible from an existing 
numerical scheme with minimal additional work.  For instance, we will see how 
OTS-NIDC makes it possible to achieve 4th-order accuracy for the diffusion 
equation in any number of space dimensions using only a second-order stencil 
for the Laplacian and explicit time integration.  
Moreover, high-order accuracy is not restricted to problems 
on simple, rectangular domains; irregular domains are straightforward to 
handle by appropriately setting the values in ghost 
cells~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
OTS-NIDC is even beneficial when using finite difference schemes to solve some 
nonlinear PDEs.  While there are certainly limitations to OTS-NIDC, we will 
see that there are many problems where it is useful.

This article is organized as follows.  In the remainder of this section,
we compare OTS-NIDC to a few related numerical techniques.
In Section~\ref{sec:OTS-NIDC}, we present the theoretical foundations of 
OTS-NIDC and show how it boosts the order of accuracy for formally low-order 
numerical schemes.  
In Section~\ref{sec:applications_1d}, we demonstrate the utility of 
OTS-NIDC to problems in one space dimension by applying it to finite 
difference schemes for several classical linear and nonlinear PDEs.  
In Section~\ref{sec:applications_multidim}, we show that 
OTS-NIDC is useful for problems in multiple space dimensions by applying it
the 2D linear advection equation and the 2D diffusion equation.  To illustrate 
the ease with which high-order accuracy can be achieved for problems on 
arbitrary domains, the 2D diffusion equation is solved on both regular 
and irregular domains.  
For each PDE in Sections~\ref{sec:applications_1d} and 
\ref{sec:applications_multidim}, we provide the numerical analysis required 
to calculate the optimal time step size, the defect correction terms, and the 
order of accuracy.  Finally, we offer some concluding remarks in 
Section~\ref{sec:summary}.

\subsection{Comparison with Traditional Defect/Deferred Correction Methods}
Defect/deferred correction is a well-known strategy for boosting the order of 
accuracy of finite difference schemes for time dependent PDEs.  It usually 
takes one of two forms: 
(1) iterative refinement of the numerical solution~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}
or 
(2) derivation of high-order compact schemes for the spatial derivative 
operator treating time derivatives as source 
terms~\cite{spotz_2001,ito_2005,heidenreich_2007}.

For iterated defect/deferred correction, the numerical solution at each 
time step is computed by iteratively adding correction terms that are 
calculated from lower accuracy iterates.  When used to improve the 
temporal order of accuracy, the iterative approach can help to ensure 
stability of the time integration scheme~\cite{kress_2006}.  Although there 
are instances where iterative defect correction has been used to eliminate 
both spatial and temporal errors~\cite{gustafsson_2002}, it is usually used to 
improve only one of these errors~\cite{pereyra_1968,kress_2002,kress_2006}.
Note that unlike iterated defect correction, OTS-NIDC requires no iteration 
during each time step because it computes correction terms using only the 
solution at the previous time step(s) (as opposed to requiring an estimate 
of the solution at the current time).

High-order compact schemes focus on eliminating spatial discretization errors 
when deriving the semi-discrete equations.  When applied to time dependent 
PDEs, a high-order compact scheme for the spatial derivative operator is 
derived treating time derivatives as source 
terms~\cite{spotz_2001,ito_2005,heidenreich_2007}.  
After deriving the high-order spatial discretization, any time differencing 
scheme may be applied to discretize the time derivatives.  While this 
approach is very general, it really only provides a mechanism for reducing 
spatial discretization error.  Temporal discretization errors are completely 
controlled by the choice of time differencing scheme.  Note that because 
high-order compact schemes for the spatial derivative operators involve spatial 
derivatives of the source term, the resulting schemes are always effectively 
implicit (even if all of the spatial derivative terms are treated explicitly).

OTS-NIDC has several advantages over both of these defect correction 
techniques.  First, OTS-NIDC is generally more computationally efficient than 
both iterated defect correction and the high-order compact scheme approach
because it requires less memory and, for many problems, admits simple 
explicit time integration schemes.  Second, OTS-NIDC inherently controls both 
spatial and temporal numerical errors at the same time.  Finally, OTS-NIDC 
does not rely solely on the addition of finite differences of solution values 
to cancel out discretization errors.  Instead, it takes advantage of 
relationships between spatial derivatives, temporal derivatives and source 
terms to eliminate the need for finite differences approximations of 
high-order derivatives of the solution.  Compared with iterated defect 
correction, OTS-NIDC makes it easier to avoid instabilities that arise from 
the presence of finite difference discretizations of high-order temporal 
derivatives and may eliminate some of the correction terms.  Comparing to the
high-order compact scheme approach, the optimization of numerical parameters 
in OTS-NIDC means that mixed partial derivatives do not need to appear in the 
fully discrete equations.

\subsection{Comparison with Adaptive Time Stepping Methods}
OTS selection is distinct and separate from traditional adaptive time stepping 
techniques~\cite{iserles_book,shampine_2005} that are used in the 
context of the method of lines~\cite{iserles_book,gko_book}.  
While both methods reduce numerical error by controlling the time step, the 
numerical errors they control are fundamentally different.  Adaptive time 
stepping can only be used to reduce temporal discretization errors because it 
controls the errors that arise when solving the coupled system of ODEs for the 
semi-discrete approximation to the PDE; the spatial discretization errors 
are completely controlled by the finite difference scheme selected to 
approximate the spatial derivatives.  
In contrast, optimal time step selection simultaneously reduces 
\emph{both} the spatial and temporal discretization errors because it
uses information from the PDE to choose the time step.  Another important 
distinction between the two techniques is that optimal time step selection 
uses a \emph{fixed} time step which makes it significantly less 
computationally complex than adaptive time stepping methods.

\section{\label{sec:OTS-NIDC}
         Theoretical Foundations of OTS Selection and NIDC}
OTS-NIDC is not by itself a method for constructing finite difference 
schemes.  Rather, it is a way to enhance to the performance of existing 
finite difference schemes by carefully choosing the time step and adding
defect correction terms to eliminate low-order numerical errors.  
The two key ideas underlying OTS-NIDC are 
(1) a judicious choice of time step can be used to eliminate the leading-order 
(and possibly higher-order) terms in the error
and
(2) the PDE provides valuable insight into the relationships between terms
in the discretization error. 
In this section, we present and analyze the technique of OTS-NIDC for finite 
difference schemes constructed to solve scalar evolution equations of the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important, 
restricting our attention to problems of this form is not a serious 
limitation.  The same principles apply to finite difference schemes for 
general PDEs.

\subsection{\label{sec:ots_model_pde} 
            Optimal Time Step for a Model PDE}
Let us begin by considering a finite difference approximation for the
following time dependent PDE in one spatial dimension: 
\beq
  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n},
  \label{eq:model_PDE}
\eeq
where $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
is well-posed.  Because we will be boosting the accuracy of the finite 
difference approximation by optimizing the time step, there is no reason to 
explicitly construct the scheme to be high-order.  In fact, using a high-order 
finite difference scheme only complicates the analysis required to calculate 
the optimal time step.

To compute the optimal time step, we first analyze the discretization errors 
(both spatial and temporal) for the finite difference approximation, 
explicitly keeping track of the leading-order terms.  In contrast to 
conventional error analysis, we do not sweep all of the errors under big-$O$
notation.  Rather, we take advantage of the fact that derivation of the 
leading-order terms in the error is straightforward for finite difference
schemes through the use of Taylor series expansions.  Because each term in 
the error is proportional to a partial derivative of the solution, we can 
use the PDE to relate the leading-order terms to each other.  
For many finite difference schemes, this careful analysis yields a direct 
relationship between the leading-order spatial and temporal errors that allows 
them to be combined as a single term of the form:
\beq
  (L u) P(\dx, \dt) ,
  \label{eq:leading_order_error_model_PDE_general}
\eeq
where $L$ is a differential operator and $P$ is a simple function of the 
numerical parameters $\dx$ and $\dt$.  By selecting the numerical parameters 
so that $P = 0$, we can completely eliminate the leading-order term in the 
discretization error.  Because time dependent PDEs are often solved by 
stepping in time, it is natural to choose the time step to be a function of 
the grid spacing.  The time step selected in this way is the
\emph{optimal time step} for the finite difference scheme.  Using the optimal
time step for time stepping yields a numerical solution with an order 
of accuracy that is \emph{higher} than the formal order of the original finite 
difference scheme.

An important class of finite difference schemes for time dependent PDEs are 
those based on the method of lines.   When the method of lines approach is 
used for (\ref{eq:model_PDE}), $P$ often takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
discretization errors, and $\alpha$ and $\beta$ are constants that 
depend on the details of the PDE and finite difference scheme.  Setting 
$P = 0$, we find that the optimal time step is given by\footnote{The $\dt = 0$ 
root of $P$ is discarded because it is not numerically meaningful.}
\beq
  \dto = \left(\alpha/\beta \right)^{1/s} \dx^{r/s}.
  \label{eq:optimal_time_step}
\eeq
Notice that the optimal time step has a similar functional dependence on the 
grid spacing as a time step that satisfies a stability constraint.

An important assumption in the above analysis is that $\alpha$ and $\beta$ 
have the same sign.  If this condition is not satisfied, (\eg finite
difference scheme for the diffusion equation based on the standard central
difference stencil for the Laplacian and backward Euler time integration),
then OTS selection is not applicable because none of the zeros of 
(\ref{eq:leading_order_error_model_PDE_MOL}) is positive.  Fortunately, 
in this situation, it is usually possible to find a closely related finite 
difference scheme which \emph{does} have a positive optimal time step.

\subsubsection*{\label{sec:error_analysis} 
            Error Analysis}
The form of the local error in 
(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
that the method of lines first approximates the PDE as a system of ODEs in 
time by only discretizing the spatial derivatives: 
\beq
{\mathbf u}_t = F({\mathbf u}),
\label{eq:method_of_lines}
\eeq
where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
$F({\mathbf u})$ is the finite difference approximation of the spatial 
derivative operator acting on ${\mathbf u}$.  The second term in the local 
error (\ref{eq:leading_order_error_model_PDE_MOL}) arises directly from 
temporal discretization errors associated with the numerical scheme used to 
integrate the system of ODEs in time.  The first term arises because $F$ is 
only an approximation of the spatial derivative operators in the PDE.  As a 
result, the system of ODEs~(\ref{eq:method_of_lines}) has an error due to 
the spatial discretization at \emph{every} point in time.  For a spatial 
discretization error of order $r$, the error in (\ref{eq:method_of_lines}) is 
$O(\dx^r)$, so the error accumulated during each time step due to 
spatial discretization error is $O(\dt \dx^r)$.  Because the method of 
lines is so frequently used to construct finite difference approximations for 
time dependent PDEs, we will focus primarily on finite difference schemes of 
this type for the remainder of the article.  It is straightforward to extend 
the results we obtain to more general finite difference schemes (\eg the
Lax-Wendroff scheme for the linear advection equation).

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First completed draft: 2008/05/12, Vitamin D, Inc.
% Submission draft: 2008/06/08, Vitamin D, Inc.
% Resubmission draft: 2008/11/16, Vitamin D, Inc.
% Resubmission draft: 2009/05/25, Serendipity Research
% Resubmission draft: 2010/03/25, Serendipity Research
% Resubmission draft: 2010/05/21, Serendipity Research
% Resubmission draft: 2011/07/30, Serendipity Research
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\documentclass[fleqn,12pt,twoside,draft]{article}
\documentclass[fleqn,12pt,twoside]{article}
\usepackage[headings]{espcrc1}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subfig}
\PassOptionsToPackage{ctagsplt,righttag}{amsmath}
\usepackage{multirow}

@@

\author{Kevin T. Chu\address{Serendipity Research, Mountain View, CA 94041, United States}
}

@@

\begin{abstract}
In this article, we present a unified analysis of the simple technique for
boosting the order of accuracy of finite difference schemes for time dependent
partial differential equations by optimally selecting the time step used to
advance the numerical solution and adding defect correction terms in a
non-iterative manner.
The power of the technique is its ability to increase the accuracy of formally
low-order finite difference schemes without major modification to the basic 
numerical algorithm.  Through straightforward numerical analysis arguments, we
explain the origin of the boost in accuracy and estimate the computational cost
of the resulting numerical method.  We demonstrate the utility of optimal time
step (OTS) selection combined with non-iterative defect correction (NIDC) on
several different types of finite difference schemes for a wide array of
classical linear and semilinear PDEs in one and more space dimensions on both
regular and irregular domains.

% KEYWORDS
\ \newline
Keywords: optimal time step; non-iterative defect correction; finite difference
schemes; high-order accurate numerical methods; time dependent PDEs

\end{abstract}

@@

\section{Introduction}
High-order numerical methods for partial differential equations (PDEs) will 
always be valuable for increasing the computational efficiency of numerical 
simulations.  Thus, it is not at all surprising that a great deal of effort in 
numerical PDEs continues to be focused on the development of high-order 
numerical schemes
\cite{spotz_2001,bruger_2005,gibou_2005,ito_2005,shukla_2005,heidenreich_2007,shukla_2007}.  
Typically, high-order accuracy is achieved by constructing
schemes that are formally high-order accurate.  However, high-order 
accuracy can also be obtained by using formally low-order schemes in clever 
ways (\eg Richardson extrapolation and compact finite difference 
schemes \cite{spotz_2001}).  When possible, the latter approach can be a 
powerful way to boost the accuracy of a numerical method without
introducing too much additional algorithmic (and programming) complexity.

One increasingly important approach for improving the accuracy of a numerical
algorithm is optimization of its numerical parameters (\eg $\dx$ and $\dt$).
For some problems, this optimization can dramatically increase the accuracy of
the numerical solution. Two well-known examples are: (1) the unit 
Courant-Friedrichs-Lewy (CFL) condition,
$\dt = \dx/a$, for the first-order upwind forward Euler discretization of the
linear advection equation, which completely eliminates the numerical error
introduced by the scheme \cite{leveque_book_2002} and (2) the boost from
second- to fourth-order accuracy that results if the time step is set to
$\dt = \dx^2/6D$ when solving the 1D diffusion equation using forward Euler
time integration with the standard second-order central difference
approximation for the Laplacian.

In this article, we present a unified analysis of the simple technique for
boosting the \emph{order} of accuracy of finite difference schemes for time
dependent PDEs by optimally choosing the time step and, when necessary, adding
defect correction terms in a \emph{non-iterative} manner.  This technique is a
systematic generalization of the ideas underlying the previously mentioned
examples and specific examples that are scattered throughout the literature.
Optimal time step (OTS) selection is based on the observation that a carefully
chosen time step can \emph{simultaneously} eliminate low-order terms in both
the spatial and temporal the discretization errors.  At its essence, OTS
selection combines the powerful idea of optimally choosing numerical parameters
with the approach of using knowledge of the structure of the PDE to reduce
numerical error.  The end result is a strategy for boosting the order of
accuracy of algorithmically simple and computationally low-cost numerical
schemes.

While OTS selection alone is sufficient to allow a formally low-order method to
deliver high-order accuracy for some problems, defect correction terms are
generally required to eliminate residual terms in the leading-order truncation
error.  However, unlike traditional defect correction 
methods \cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}, 
using OTS selection eliminates the need to incorporate correction terms that
involve derivatives of higher-order than those that appear in the original PDE.
Another significant feature of OTS selection with defect correction
is that the defect terms are incorporated in a \emph{non-iterative} fashion.
In contrast, traditional defect/deferred correction approaches typically
require several refinement iterations at each time step to boost the order of
accuracy of the numerical solution.  OTS with non-iterative defect correction
(OTS-NIDC) also differs from the method of modified
equations \cite{griffiths_1986} for eliminating higher-order
corrections to the truncation error because it uses an optimal time
step, not explicit inclusion of correction terms, to regularize the
singularly perturbed modified equations - the correction terms that remain
when using OTS-NIDC are all \emph{regular} perturbations to the original
equation.

As a technique for enhancing numerical methods, OTS-NIDC has several desirable
features.  Like any technique that leads to high-order methods, OTS-NIDC
produces schemes that greatly reduce the computational cost required to obtain
a numerical solution of a desired accuracy.  However, its real power comes
from the fact that it achieves high-order accuracy while being based solely on
very simple, formally low-order finite difference schemes \emph{without}
requiring special algorithmic steps.  In other words, OTS-NIDC allows us to
extract as much accuracy as possible from simple numerical schemes with
minimal additional work.  For instance, we will see how OTS-NIDC makes it
possible to achieve 4th-order accuracy for the diffusion equation in any
number of space dimensions using only a second-order stencil for the Laplacian
and explicit time integration.  Moreover, high-order accuracy is not
restricted to problems on simple, rectangular domains; irregular domains are
straightforward to handle by appropriately setting the values in ghost
cells \cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
OTS-NIDC is even beneficial when using finite difference schemes to solve
some nonlinear PDEs.

This article is organized as follows.
In Section~\ref{sec:theoretical_comparisons}, we compare OTS-NIDC to 
related numerical techniques.
In Section~\ref{sec:OTS-NIDC}, we present the theoretical foundations of 
OTS-NIDC and show how it boosts the order of accuracy for formally
low-order numerical schemes.  
In Section~\ref{sec:applications_1d}, we demonstrate the utility of 
OTS-NIDC to problems in one space dimension by applying it to finite 
difference schemes for several classical linear and nonlinear PDEs.  
In Section~\ref{sec:applications_multidim}, we show that 
OTS-NIDC is useful for problems in multiple space dimensions by applying
it to the 2D linear advection equation and the 2D diffusion equation.  To
illustrate the ease with which high-order accuracy can be achieved for
problems on arbitrary domains, the 2D diffusion equation is solved on both
regular and irregular domains.  For each PDE in
Sections~\ref{sec:applications_1d} and \ref{sec:applications_multidim}, we
provide the numerical analysis required to calculate the optimal time step
size, the defect correction terms, and the order of accuracy.  Note that in
order to emphasize the method, we have chosen to apply it to relatively simple
problems rather than use it to tackle ``real-life'' applications.  The method,
however, is not restricted to these simple examples.  Finally, 
Section~\ref{sec:conclusions} offers some concluding remarks.

\section{Theoretical Comparisons with Related Numerical Techniques
         \label{sec:theoretical_comparisons}}

\subsection{Comparison with Defect/Deferred Correction Methods}
Defect/deferred correction is a well-known strategy for boosting the order of 
accuracy of finite difference schemes for time dependent PDEs.  It usually 
takes one of two forms: 
(1) iterative refinement of the numerical
solution \cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}
or 
(2) derivation of high-order compact schemes for the spatial derivative 
operator treating time derivatives as source 
terms \cite{spotz_2001,ito_2005,heidenreich_2007}.

For iterated defect/deferred correction, the numerical solution at each 
time step is computed by iteratively adding correction terms that are 
calculated from lower accuracy iterates.  When used to improve the 
temporal order of accuracy, the iterative approach helps to ensure 
stability of the time integration scheme \cite{kress_2006}.  Although there 
are instances where iterative defect correction has been used to eliminate 
both spatial and temporal errors \cite{gustafsson_2002}, it is usually used to 
improve only one of these errors \cite{pereyra_1968,kress_2002,kress_2006}.
Unlike iterated defect correction, OTS-NIDC requires no iteration during each
time step because it computes correction terms using only the solution at the
previous time step(s) (as opposed to requiring an estimate of the solution at
the current time).

High-order compact schemes focus on eliminating spatial discretization errors 
when deriving the semi-discrete equations.  When applied to time dependent 
PDEs, a high-order compact scheme for the spatial derivative operator is 
derived treating time derivatives as source 
terms \cite{spotz_2001,ito_2005,heidenreich_2007}.  
After deriving the high-order spatial discretization, any time differencing 
scheme may be applied to discretize the time derivatives.  While this 
approach is very general, it really only provides a mechanism for reducing 
spatial discretization error.  Temporal discretization errors are completely 
controlled by the choice of time differencing scheme.  Note that because 
high-order compact schemes for the spatial derivative operators involve spatial 
derivatives of the source term, the resulting schemes are always effectively 
implicit (even if all of the spatial derivative terms are treated explicitly).

OTS-NIDC has several advantages over both of these defect correction 
techniques.  First, OTS-NIDC is generally more computationally efficient than 
both iterated defect correction and the high-order compact scheme approach
because it requires less memory and, for many problems, admits simple 
explicit time integration schemes.  Second, OTS-NIDC inherently controls both 
spatial and temporal numerical errors at the same time.  Finally, taking
advantage of relationships between spatial derivatives, temporal derivatives
and source terms, OTS-NIDC avoids the need to explicitly incorporate finite
differences of high-order derivatives of intermediate solutions in order to
eliminate discretization errors. 
Compared with iterated defect correction, OTS-NIDC makes it easier to avoid
instabilities that arise from the presence of finite difference discretizations
of high-order temporal derivatives by automatically eliminating the most
problematic correction terms.  Comparing to the high-order compact scheme
approach, the use of an optimal time step in OTS-NIDC means that mixed 
spatio-temporal partial derivatives do not need to appear in the fully
discrete equations.

\subsection{Comparison with the Modified Equation Method}
Both defect and deferred correction can be viewed as special cases of the
method of modified equations \cite{griffiths_1986}.  When using any modified
equation method, the basic approach is to derive a modified version of the 
original PDE through the introduction of the leading-order terms of the
discretization error into the PDE.  Unfortunately, the modified PDE is a
singular perturbation to the original PDE due to the presence of the
discretization parameters in the correction terms.  Because the the singular
perturbation can lead to numerical instabilities, finding a way regularize
the modified equation is an important step for developing any numerical method
based on the modified equations method.

Whereas deferred correction methods deal with the singular perturbation by
using an iterative procedure to avoid numerical instabilities, the traditional
method of modified equations \cite{griffiths_1986} handles the
singularity by ``pushing'' the singularity to a higher-order term in the
discretization error.  The basic procedure involves substituting derivatives
that appear in the singular term with expressions obtained by taking
derivatives of the original PDE.  This procedure yields a new modified
equation where high-order derivatives of the solution are replaced by
derivatives of source terms and the new singular term is of sufficiently
high-order in the discretization parameter that it can be neglected in the
construction of the numerical scheme.

While OTS-NIDC, like defect/deferred correction, falls within the general
framework of the method of modified equations, it is an important special
case because regularization of the singular perturbation in the modified
equations is \emph{automatic}.  Regularization results from selecting the time
step so that the spatial and temporal singular perturbation terms exactly
cancel each other rather than through an iterative refinement procedure.
By constraining the relationship between the time step and grid spacing, the
modified equation becomes a regular perturbation to the original PDE.  In the
special case when OTS-NIDC does not require defect correction, the modified
equation is actually exactly the same as the original equation -- no regular
perturbation terms remain after the singular perturbation terms have been
eliminated.

There are several advantages of OTS-NIDC over traditional modified
equation methods.  First, because no finite difference approximations of
higher-order derivatives are required, numerical schemes based on OTS-NIDC
do not require additional boundary conditions.  Second, derivation
of the optimal time step and correction terms for OTS-NIDC requires a comparable
amount of analysis as the traditional modified equation method.  In both cases,
we need explicit knowledge of the leading-order terms in the discretization
error.  Finally, because the leading-order terms in the discretization error
are eliminated implicitly rather than explicitly as additional terms in the 
numerical scheme, the computational performance of OTS-NIDC is often
superior to the performance of traditional modified equation methods.

\subsection{Comparison with Adaptive Time Stepping Methods}
OTS-NIDC is distinct and separate from traditional adaptive time stepping 
techniques \cite{iserles_book,shampine_2005} that are used in the 
context of the method of lines \cite{iserles_book,gko_book}.  
While both methods reduce numerical error by controlling the time step, the 
numerical errors they control are fundamentally different.  Adaptive time 
stepping can only be used to reduce temporal discretization errors because it 
controls the errors that arise when solving the coupled system of ODEs for the 
semi-discrete approximation to the PDE; the spatial discretization errors 
are completely controlled by the finite difference scheme selected to 
approximate the spatial derivatives.  
In contrast, optimal time step selection simultaneously reduces 
\emph{both} the spatial and temporal discretization errors because it
uses information from the PDE to choose the time step.  Another important 
distinction between the two techniques is that optimal time step selection 
uses a \emph{fixed} time step which makes it significantly less 
computationally complex than adaptive time stepping methods.

\section{\label{sec:OTS-NIDC} Theoretical Foundations of OTS-NIDC}
Technically, OTS-NIDC is not by itself a method for constructing finite
difference schemes.  Rather, it is a way to enhance to the performance of
existing finite difference schemes by carefully choosing the time step to
eliminate leading-order terms in the discretization error.  The two key ideas
underlying OTS-NIDC are 
(1) a judicious choice of time step can be used to eliminate the leading-order 
(and possibly higher-order) terms in the error
and
(2) the PDE provides valuable insight into the relationships between terms
in the discretization error. 
In this section, we present and analyze the technique of OTS-NIDC for
finite difference schemes constructed to solve scalar evolution equations of
the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important, 
restricting our attention to problems of this form is not a serious 
limitation.  The same principles apply to finite difference schemes for 
general PDEs.

\subsection{\label{sec:ots_model_pde} 
            Optimal Time Step Selection for a Model PDE}
Let us begin by considering a finite difference approximation for the
following time dependent PDE in one spatial dimension: 
\beq
  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n},
  \label{eq:model_PDE}
\eeq
where $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
is well-posed.  Because we will be boosting the accuracy of the finite 
difference approximation by optimizing the time step, there is no reason to 
explicitly construct the scheme to be high-order.  In fact, using a high-order 
finite difference scheme only complicates the analysis required to calculate 
the optimal time step.

To compute the optimal time step, we first analyze the discretization errors 
for the finite difference approximation, explicitly keeping track of the
leading-order terms.  Because each term in the error is proportional to a
partial derivative of the solution, we can use the PDE to relate the
leading-order terms to each other.  For many finite difference schemes, this
careful analysis yields a direct relationship between the leading-order spatial
and temporal errors that allows them to be combined as a single term of the
form:
\beq
  (L u) P(\dx, \dt) ,
  \label{eq:leading_order_error_model_PDE_general}
\eeq
where $L$ is a differential operator and $P$ is a simple function of the 
numerical parameters $\dx$ and $\dt$.  By selecting the time step
so that $P = 0$, we can completely eliminate the leading-order term in the 
discretization error.  The time step selected in this way is the
\emph{optimal time step} for the finite difference scheme.  Using the optimal
time step for time stepping yields a numerical solution with an
\emph{effective} order of accuracy that is higher than the formal order of the
original finite difference scheme.

An important class of finite difference schemes for time dependent PDEs are 
those based on the method of lines.   When the method of 
lines \cite{iserles_book,gko_book,cash_1996,psihoyios_2007}.  
approach is 
used for (\ref{eq:model_PDE}), $P$ often takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
discretization errors, and $\alpha$ and $\beta$ are constants that 
depend on the details of the PDE and finite difference scheme.  Setting 
$P = 0$, we find that the optimal time step is given by\footnote{The $\dt = 0$ 
root of $P$ is discarded because it is not numerically meaningful.}
\beq
  \dto = \left(\alpha/\beta \right)^{1/s} \dx^{r/s}.
  \label{eq:optimal_time_step}
\eeq

An important assumption in the above analysis is that $\alpha$ and $\beta$ 
have the same sign.  If this condition is not satisfied, (\eg finite
difference scheme for the diffusion equation based on the standard central
difference stencil for the Laplacian and backward Euler time integration),
then OTS selection is not applicable because none of the zeros of 
(\ref{eq:leading_order_error_model_PDE_MOL}) is positive.  Fortunately, 
in this situation, it is usually possible to find a closely related finite 
difference scheme that has a positive optimal time step.
